\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2021-07-09 submission/manuscript.tex
Review done.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{times} % Times New Roman font
\usepackage[T1]{fontenc}

@@

\usepackage[labelsep=period, labelfont=bf]{caption}
\renewcommand{\thefigure}{\arabic{figure}}
\renewcommand{\figurename}{Fig.}
\captionsetup{justification=raggedright,singlelinecheck=false}

@@

\renewcommand{\figureautorefname}{Fig.}

@@

@@

3. NIOZ, Department of Marine Microbiology and Biogeochemistry, Royal
Netherlands Institute for Sea Research, Utrecht University, The
Netherlands

@@

Surfaces of marine macrophytes are inhabited by diverse microbial
communities. Most studies focusing on epiphytic communities of
macrophytes did not take into account temporal changes or applied low
sampling frequency approaches. The seasonal dynamics of epiphytic
microbial communities was determined in a meadow of \emph{Cymodocea
nodosa} invaded by \emph{Caulerpa cylindracea} and in a monospecific
settlement of \emph{Caulerpa cylindracea} at monthly intervals. For
comparison the ambient prokaryotic picoplankton community was also
characterized. At the OTU level, the microbial community composition
differed between the ambient water and the epiphytic communities
exhibiting host-specificity. Also, successional changes were observed
connected to the macrophyte growth cycle. Taxonomic analysis, however,
showed similar high rank groups in the ambient water and the epiphytic
communities, with the exception of \emph{Desulfobacterota}, which were
only found on \emph{Caulerpa cylindracea}. \emph{Cyanobacteria} showed
seasonal changes while other high rank taxa were present throughout the
year. Phylogenetic groups present throughout the year constituted most
of the sequences, while less abundant taxa showed seasonal patterns
connected to the macrophyte growth cycle. Taken together, epiphytic
microbial communities of the seagrass \emph{Cymodocea nodosa} and the
macroalga \emph{Caulerpa cylindracea} appear to be host-specific and
contain taxa that undergo successional changes.

@@

Marine macrophytes (seagrasses and macroalgae) are important ecosystem
engineers forming close associations with microorganisms belonging to
all three domains of life (Egan et al., 2013; Tarquinio et al., 2019).
Microbes can live within macrophyte tissue as endophytes or can form
epiphytic communities on surfaces of leaves and thalli (Egan et al.,
2013; Hollants et al., 2013; Tarquinio et al., 2019). Epiphytic and
endophytic microbial communities exhibit a close functional relationship
with the macrophyte host. It was proposed that this close relationship
constitutes a holobiont, an integrated community where the macrophyte
organism and its symbiotic partners support each other (Margulis, 1991;
Egan et al., 2013; Tarquinio et al., 2019).

Biofilms of microbial epiphytes can contain diverse taxonomic groups and
harbor cell abundances from 10\textsuperscript{2} to
10\textsuperscript{7} \si{\cells\per\cm\squared} (Armstrong et al.,
2000; Bengtsson et al., 2010; Burke et al., 2011b). In such an
environment a number of positive and negative interactions between the
macrophyte and the colonizing microorganisms have been described (Egan
et al., 2013; Hollants et al., 2013; Tarquinio et al., 2019).
Macrophytes can promote growth of associated microbes by nutrient
exudation, while in return microorganisms may support macrophyte
performance through improved nutrient availability, phytohormone
production and protection from toxic compounds, oxidative stress,
biofouling organisms and pathogens (Egan et al., 2013; Hollants et al.,
2013; Tarquinio et al., 2019). Besides these positive interactions,
macrophytes can negatively impact the associated microbes such as
pathogenic bacteria by producing reactive oxygen species and secondary
metabolites (Egan et al., 2013; Hollants et al., 2013; Tarquinio et al.,
2019).

All these ecological roles are carried out by a taxonomically diverse
community of microorganisms. At higher taxonomic ranks a core set of
macrophyte epiphytic taxa was described consisting mainly of
\emph{Alphaproteobacteria}, \emph{Gammaproteobacteria},
\emph{Desulfobacterota}, \emph{Bacteroidota}, \emph{Cyanobacteria},
\emph{Actinobacteriota}, \emph{Firmicutes}, \emph{Planctomycetota},
\emph{Chloroflexi} and \emph{Verrucomicrobiota} (Crump and Koch, 2008;
Tujula et al., 2010; Lachnit et al., 2011). In contrast, at lower
taxonomic ranks host specific microbial communities were found (Lachnit
et al., 2011; Roth-Schulze et al., 2016). Recently, it was shown that
even different morphological niches within the same alga had a higher
influence on the composition of the bacterial community than
biogeography or environmental factors (Morrissey et al., 2019). While
the microbial community composition varies between host species,
metagenomic analyses revealed that the majority of the microbial
functions are conserved (Burke et al., 2011a; Roth-Schulze et al.,
2016). This discrepancy between the microbial taxonomic and functional
composition might be explained by the lottery hypothesis. It postulates
that an initial random colonization step takes places from a set of
functionally equivalent taxonomic groups resulting in taxonomically
different epiphytic communities sharing a core set of functional genes
(Burke et al., 2011a; Roth-Schulze et al., 2016). In addition, some of
the variation in the reported data could be attributed to different
techniques used in these studies, such as different protocols for
epiphytic cell detachment and/or DNA isolation, as no standard protocol
has been yet established to study epiphytic communities (Ugarelli et
al., 2019; Korlević et al., submitted).

The majority of studies describing macrophyte epiphytic microbial
communities did not include possible seasonal changes (Crump and Koch,
2008; Lachnit et al., 2009; Burke et al., 2011b; Roth-Schulze et al.,
2016; Ugarelli et al., 2019). If seasonal changes were taken into
account, low temporal frequency, applied methodologies and/or limited
number of analysed host species did not allow a high taxonomic
resolution (Tujula et al., 2010; Lachnit et al., 2011; Michelou et al.,
2013; Miranda et al., 2013; Mancuso et al., 2016). In the present study
we describe the seasonal dynamics of bacterial and archaeal communities
on the surfaces of the seagrass \emph{Cymodocea nodosa} and siphonous
macroalgae \emph{Caulerpa cylindracea} determined on a mostly monthly
scale. Bacterial and archaeal epiphytes were sampled in a meadow of
\emph{C. nodosa} invaded by the invasive \emph{C. cylindracea} and in a
locality of only \emph{C. cylindracea} located in the proximity of the
seagrass meadow. For comparison, the microbial community of the ambient
seawater was also characterized.

@@

Sampling was performed in the Bay of Funtana, northern Adriatic Sea
(\ang{45;10;39} N, \ang{13;35;42} E). Leaves and thalli were sampled in
a meadow of \emph{Cymodocea nodosa} invaded by the invasive
\emph{Caulerpa cylindracea} (mixed settlement) and in a monospecific
settlement of \emph{Caulerpa cylindracea} located in the proximity of
the meadow at approximately monthly intervals from December 2017 to
October 2018 (\autoref{supp-nseq_notus}). Leaves and thalli were
collected by diving and transported to the laboratory in containers
placed on ice and filled with seawater collected at the sampling site.
Upon arrival to the laboratory, \emph{C. nodosa} leaves were cut into
sections of 1 -- 2 \si{\cm}, while \emph{C. cylindracea} thalli were cut
into 5 -- 8 \si{\cm} long sections. Leaves and thalli were washed three
times with sterile artificial seawater (ASW) to remove loosely attached
microbial cells. Ambient seawater was collected in 10 \si{\l} containers
by diving and transported to the laboratory where 10 -- 20 \si{\l} were
filtered through a 20 \si{\um} net. The filtrate was further
sequentially filtered through 3 \si{\um} and 0.2 \si{\um} polycarbonate
membrane filters (Whatman, United Kingdom) using a peristaltic pump.
Filters were briefly dried at room temperature and stored at \num{-80}
\si{\degreeCelsius}. Seawater samples were also collected approximately
monthly from July 2017 to October 2018.

@@

DNA from surfaces of \emph{C. nodosa} and \emph{C. cylindracea} was
isolated using a previously modified and adapted protocol that allows
for a selective epiphytic DNA isolation (Massana et al., 1997; Korlević
et al., submitted). Briefly, leaves and thalli are incubated in a lysis
buffer and treated with lysozyme and proteinase K. Following the
incubations, the mixture containing lysed epiphytic cells was separated
from the leaves and thalli and extracted using phenol-chloroform.
Finally, the extracted DNA was precipitated using isopropanol. DNA from
seawater picoplankton was extracted from 0.2 \si{\um} polycarbonate
filters according to Massana et al. (1997) with a slight modification.
Following the phenol-chloroform extraction, 1/10 of chilled 3
\si{\Molar} sodium acetate (pH 5.2) was added. DNA was precipitated by
adding 1 volume of chilled isopropanol, incubating the mixtures
overnight at \num{-20} \si{\degreeCelsius} and centrifuging at 20,000 ×
\emph{g} and 4 \si{\degreeCelsius} for 21 \si{\minute}. The pellet was
washed twice with 500 \si{\ul} of chilled 70 \si{\percent} ethanol and
centrifuged after each washing step at 20,000 × \emph{g} and 4
\si{\degreeCelsius} for 5 \si{\minute}. Dried pellets were re-suspended
in 50 -- 100 \si{\ul} of deionized water.

@@

Illumina MiSeq sequencing of the V4 region of the 16S rRNA gene was
performed as described previously (Korlević et al., submitted). The V4
region of the 16S rRNA gene was amplified using a two-step PCR
procedure. In the first PCR, the 515F
(\(5'\)-GTGYCAGCMGCCGCGGTAA-\(3'\)) and 806R
(\(5'\)-GGACTACNVGGGTWTCTAAT-\(3'\)) primers from the Earth Microbiome
Project (\url{https://earthmicrobiome.org/protocols-and-standards/16s/})
were used (Caporaso et al., 2012; Apprill et al., 2015; Parada et al.,
2016). These primers contained on their \(5'\) end a tagged sequence.
Purified PCR products were sent for Illumina MiSeq sequencing at IMGM
Laboratories, Martinsried, Germany. Prior to sequencing at IMGM, the
second PCR amplification of the two-step PCR procedure was performed
using primers targeting the tagged region incorporated in the first PCR.
In addition, these primers contained adapter and sample-specific index
sequences. Beside samples, a positive and negative control for each
sequencing batch was sequenced. The negative control comprised PCR
reactions without DNA template, while for a positive control a mock
community composed of evenly mixed DNA material originating from 20
bacterial strains (ATCC MSA-1002, ATCC, USA) was used. Sequences
obtained in this study have been deposited in the European Nucleotide
Archive (ENA) at EMBL-EBI under accession number PRJEB37267.

\hypertarget{sequence-analysis}{%
\subsubsection{Sequence analysis}\label{sequence-analysis}}

Obtained sequences were analysed on the computer cluster Isabella
(University Computing Center, University of Zagreb) using mothur
(version 1.43.0) (Schloss et al., 2009) according to the MiSeq Standard
Operating Procedure (MiSeq SOP; \url{https://mothur.org/wiki/MiSeq_SOP})
(Kozich et al., 2013) and recommendations provided by the Riffomonas
project to enhance data reproducibility
(\url{http://www.riffomonas.org/}). For alignment and classification of
sequences the SILVA SSU Ref NR 99 database (release 138;
\url{http://www.arb-silva.de}) was used (Quast et al., 2013; Yilmaz et
al., 2014). Pipeline data processing and visualization was done using R
(version 3.6.0) (R Core Team, 2019) combined with packages vegan
(version 2.5-6) (Oksanen et al., 2019), tidyverse (version 1.3.0)
(Wickham et al., 2019) and multiple other packages (Neuwirth, 2014; Xie,
2014, 2015, 2019a, 2019b; Wilke, 2018; Xie et al., 2018; Allaire et al.,
2019; Zhu, 2019). The detailed analysis procedure including the R
Markdown file are available in the GitHub repository
(\url{https://github.com/MicrobesRovinj/Korlevic_EpiphyticDynamics_FrontMicrobiol_2021}).
Based on the ATCC MSA-1002 mock community included in the analysis an
average sequencing error rate of 0.01 \si{\percent} was determined,
which is in line with previously reported values for next-generation
sequencing data (Kozich et al., 2013; Schloss et al., 2016). In
addition, the negative controls processed together with the samples
yielded on average only 2 sequences after sequence quality curation.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{times} % Times New Roman font
\usepackage[T1]{fontenc}

@@

\usepackage[labelsep=period, labelfont=bf]{caption}
\renewcommand{\thefigure}{\arabic{figure}}
\renewcommand{\figurename}{Figure}
\captionsetup{justification=raggedright,singlelinecheck=false}

@@

\renewcommand{\figureautorefname}{Figure}

@@

\DeclareSIUnit{\litre}{l}
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

3. Department of Marine Microbiology and Biogeochemistry, Royal
Netherlands Institute for Sea Research (NIOZ), Utrecht University, The
Netherlands

@@

Surfaces of marine macrophytes are inhabited by diverse microbial
communities. Most studies focusing on epiphytic communities of
macrophytes did not take into account temporal changes or applied low
sampling frequency approaches. The seasonal dynamics of epiphytic
microbial communities was determined in a meadow of \emph{Cymodocea
nodosa} invaded by \emph{Caulerpa cylindracea} and in a monospecific
settlement of \emph{Caulerpa cylindracea} at monthly intervals. For
comparison the ambient prokaryotic picoplankton community was also
characterized. At the OTU level, the microbial community composition
differed between the ambient water and the epiphytic communities
exhibiting host-specificity. Also, successional changes were observed
connected to the macrophyte growth cycle. Taxonomic analysis, however,
showed similar high rank taxa (phyla and classes) in the ambient water
and the epiphytic communities, with the exception of
\emph{Desulfobacterota}, which were only found on \emph{Caulerpa
cylindracea}. \emph{Cyanobacteria} showed seasonal changes while other
high rank taxa were present throughout the year. In months of high
\emph{Cyanobacteria} presence the majority of cyanobacterial sequences
were classified as \emph{Pleurocapsa}. Phylogenetic groups present
throughout the year (e.g.~\emph{Saprospiraceae},
\emph{Rhodobacteraceae}, members without known relatives within
\emph{Gammaproteobacteria}, \emph{Desulfatitalea} and members without
known relatives within \emph{Desulfocapsaceae}) constituted most of the
sequences, while less abundant taxa showed seasonal patterns connected
to the macrophyte growth cycle. Taken together, epiphytic microbial
communities of the seagrass \emph{Cymodocea nodosa} and the macroalga
\emph{Caulerpa cylindracea} appear to be host-specific and contain taxa
that undergo successional changes.

@@

Marine macrophytes (seagrasses and macroalgae) are important ecosystem
engineers forming close associations with microorganisms belonging to
all three domains of life (Egan et al., 2013; Tarquinio et al., 2019).
Microbes can live within macrophyte tissue as endophytes or form
epiphytic communities on surfaces of leaves, roots, rhizomes and thalli
(Egan et al., 2013; Hollants et al., 2013; Aires et al., 2015; Tarquinio
et al., 2019). Epiphytic and endophytic microbial communities exhibit a
close functional relationship with the macrophyte host. It has been
proposed that this close relationship constitutes a holobiont, an
integrated community where the macrophyte organism and its symbiotic
partners support each other (Margulis, 1991). In addition, as suggested
by the hologenome theory endophytic microbes play a critical role in the
adaptation and evolution of the host species (Aires et al., 2015).

Biofilms of microbial epiphytes can contain diverse taxonomic groups and
harbour cell abundances from 10\textsuperscript{2} to
10\textsuperscript{7} \si{\cells\per\cm\squared} (Armstrong et al.,
2000; Bengtsson et al., 2010; Burke et al., 2011b). In such an
environment a number of positive and negative interactions between the
macrophyte and the colonizing microorganisms have been described (Egan
et al., 2013; Hollants et al., 2013; Tarquinio et al., 2019).
Macrophytes can promote growth of associated microbes by nutrient
exudation (Wood and Hayasaka, 1981), while in return microorganisms may
support macrophyte performance through improved nutrient availability
(Nielsen et al., 2001; de Oliveira et al., 2012), phytohormone
production (Matsuo et al., 2003; Celdrán et al., 2012) and protection
from toxic compounds (Küsel et al., 2006), oxidative stress
(Sanchez-Amat et al., 2010), biofouling organisms (Dobretsov and Qian,
2002) and pathogens (Penesyan et al., 2009). Besides these positive
interactions, macrophytes can negatively impact the associated microbes
by producing reactive oxygen species (Weinberger, 2007) and secondary
metabolites (Saha et al., 2011).

All these ecological roles are carried out by a taxonomically diverse
community of microorganisms. At higher taxonomic ranks (phyla and
classes) microbial taxa, such as \emph{Alphaproteobacteria},
\emph{Gammaproteobacteria}, \emph{Bacteroidota} and
\emph{Cyanobacteria}, have been associated with surfaces of seagrass
leaves and macroalgal thalli (Crump and Koch, 2008; Tujula et al., 2010;
Lachnit et al., 2011; Egan et al., 2013; Tarquinio et al., 2019;
Ugarelli et al., 2019). While similar high rank taxa have been found on
surfaces of different macrophyte species, in order to describe new
ecological patterns it is also necessary to focus on lower taxonomic
ranks (genus and OTUs) which tend to be host-specific (Lachnit et al.,
2011; Hollants et al., 2013; Roth-Schulze et al., 2016). While the
microbial community composition can vary between host species,
metagenomic analyses revealed that the majority of microbial functions
are conserved, showing that different epiphytic microbial species could
be functionally similar (Burke et al., 2011a; Roth-Schulze et al., 2016;
Cúcio et al., 2018). This discrepancy between the microbial taxonomic
and functional composition might be explained by the lottery hypothesis
(Sale, 1976). It postulates that an initial random colonization step
takes place from a set of functionally equivalent taxonomic groups
resulting in taxonomically different epiphytic communities sharing a
core set of functional genes (Burke et al., 2011a; Stratil et al., 2013;
Schmidt et al., 2015; Roth-Schulze et al., 2016).

Seagrasses are know to form close relationships with microbial
communities associated with the surfaces of leaves, roots and rhizomes
(Cúcio et al., 2016; Crump et al., 2018; Ugarelli et al., 2019; Ettinger
and Eisen, 2020; Wang et al., 2020). For different seagrass species a
distinct microbial community from ambient seawater or bulk sediment has
been reported, however no species specific communities have been found
(Cúcio et al., 2016; Crump et al., 2018; Ugarelli et al., 2019). It
seems that seagrasses are selecting the associated microbial community
but these microbes have not coevolved with their seagrass plant host.
Similar to seagrasses, siphonous macroalgae of the genus \emph{Caulerpa}
are also closely associated with their microbial communities (Aires et
al., 2013, 2015; Rizzo et al., 2016b; Stabili et al., 2017; Morrissey et
al., 2019). While some studies have found similar culturable bacterial
groups associated with the surface of a \emph{Caulerpa} species from
different geographic locations (Stabili et al., 2017), other have
reported large compositional differences that were mainly attributed to
different host species of this genus, biogeography and nutrient levels
(Morrissey et al., 2019) again raising the question to which extent are
associated communities host-specific.

Since marine macrophytes in temperate zones are exhibiting seasonal
changes in growth and physiology (Agostini et al., 2003; Najdek et al.,
2020) it is important to verify if and how surface associated microbial
communities are affected by these changes. The majority of studies
describing macrophyte epiphytic microbial communities have not included
possible seasonal changes (Crump and Koch, 2008; Lachnit et al., 2009;
Burke et al., 2011b; Roth-Schulze et al., 2016; Ugarelli et al., 2019).
If seasonal changes have been taken into account, low temporal
frequency, applied methodologies and/or limited number of analysed host
species did not allow for a detailed taxonomic analysis (Tujula et al.,
2010; Lachnit et al., 2011; Bengtsson et al., 2012; Michelou et al.,
2013; Miranda et al., 2013; Mancuso et al., 2016). In the present study
we performed a descriptive analysis of seasonal bacterial and archaeal
community dynamics on the surfaces of the seagrass \emph{Cymodocea
nodosa}, an abundant seagrass species in the Mediterranean (Short et
al., 2001), and siphonous macroalga \emph{Caulerpa cylindracea}, one of
the most invasive macroalgal species (Klein and Verlaque, 2008;
Boudouresque et al., 2009). Bacterial and archaeal epiphytes were
sampled in a meadow of \emph{C. nodosa} invaded by the invasive \emph{C.
cylindracea} and in a locality of only \emph{C. cylindracea} located in
the proximity of the seagrass meadow. For comparison, the microbial
community of the ambient seawater was also characterized. The presence
of both macrophytes in the same area enabled (i) the assessment of
differences in the bacterial and archaeal communities between host
species and settlements of \emph{C. cylindracea} and (ii) the evaluation
of differences between surface associated and free living (ambient
seawater) communities. In addition, these differences were evaluated on
a monthly scale providing insight into seasonal changes (iii).

@@

Sampling was performed in the Bay of Funtana, northern Adriatic Sea
(\ang{45;10;39} N, \ang{13;35;42} E). The sea floor in the bay is partly
covered by the invasive macroalga \emph{C. cylindracea} that can be
found in a monospecific settlement or mixed with the seagrass \emph{C.
nodosa} (\autoref{map}). \emph{C. nodosa} leaves were retrieved from a
meadow of \emph{C. nodosa} invaded by the invasive \emph{C. cylindracea}
(mixed settlement; depth, 2 -- 2.5 \si{\m}), while \emph{C. cylindracea}
thalli were sampled in the same invaded meadow (mixed settlement; depth,
2 -- 2.5 \si{\m}) and in a monospecific settlement (depth, 1 -- 1.5
\si{\m}) of \emph{C. cylindracea} located in the proximity (20 -- 50
\si{\m}) of the invaded meadow at approximately monthly intervals from
November 2017 to October 2018 (\autoref{supp-nseq_notus}). Leaves and
thalli were collected by diving and transported to the laboratory in
containers placed on ice and filled with seawater collected at the
sampling site. Upon arrival to the laboratory, \emph{C. nodosa} leaves
were cut into sections of 1 -- 2 \si{\cm}, while \emph{C. cylindracea}
thalli were cut into 5 -- 8 \si{\cm} long sections. Leaves and thalli
were washed three times with sterile artificial seawater (ASW) to remove
loosely attached microbial cells. Ambient seawater was collected in 10
\si{\l} containers by diving and transported to the laboratory where 10
-- 20 \si{\l} were filtered through a 20 \si{\um} net. The filtrate was
further sequentially filtered through 3 \si{\um} and 0.2 \si{\um}
polycarbonate membrane filters (Whatman, United Kingdom) using a
peristaltic pump. Filters were briefly dried at room temperature and
stored at \num{-80} \si{\degreeCelsius}. Seawater samples were also
collected approximately monthly from July 2017 to October 2018.

@@

DNA from surfaces of \emph{C. nodosa} and \emph{C. cylindracea} was
isolated from a pool of leaves or thalli on the sampling day using a
previously modified and adapted protocol that allows for a selective
epiphytic DNA isolation (Massana et al., 1997; Korlević et al., 2021).
Briefly, leaves and thalli were incubated in a lysis buffer and treated
with lysozyme and proteinase K. Following the incubations, the mixture
containing lysed epiphytic cells was separated from the leaves and
thalli and extracted using phenol-chloroform. Finally, the extracted DNA
was precipitated using isopropanol. DNA from seawater picoplankton was
extracted from 0.2 \si{\um} polycarbonate filters according to Massana
et al. (1997) with a slight modification. Following the
phenol-chloroform extraction, 1/10 of 3 \si{\Molar} sodium acetate (pH
5.2) was added. DNA was precipitated by adding 1 volume of chilled
isopropanol, incubating the mixtures overnight at \num{-20}
\si{\degreeCelsius} and centrifuging at 20,000 × \emph{g} and 4
\si{\degreeCelsius} for 20 \si{\minute}. The pellet was washed twice
with 500 \si{\ul} of chilled 70 \si{\percent} ethanol and centrifuged
after each washing step at 20,000 × \emph{g} and 4 \si{\degreeCelsius}
for 5 \si{\minute}. Dried pellets were re-suspended in 50 -- 100
\si{\ul} of deionized water.

@@

Illumina MiSeq sequencing of the V4 region of the 16S rRNA gene was
performed as described previously (Korlević et al., 2021). The V4 region
of the 16S rRNA gene was amplified using a two-step PCR procedure. In
the first PCR, the 515F (\(5'\)-GTGYCAGCMGCCGCGGTAA-\(3'\)) and 806R
(\(5'\)-GGACTACNVGGGTWTCTAAT-\(3'\)) primers from the Earth Microbiome
Project (\url{https://earthmicrobiome.org/protocols-and-standards/16s/})
were used (Caporaso et al., 2012; Apprill et al., 2015; Parada et al.,
2016). These primers contained on their \(5'\) end a tagged sequence.
Purified PCR products were sent for Illumina MiSeq sequencing at IMGM
Laboratories, Martinsried, Germany. Prior to sequencing at IMGM, the
second PCR amplification of the two-step PCR procedure was performed
using primers targeting the tagged region incorporated in the first PCR.
In addition, these primers contained adapter and sample-specific index
sequences. Beside samples, a positive and negative control for each
sequencing batch was sequenced. The negative control comprised PCR
reactions without DNA template, while for a positive control a mock
community composed of evenly mixed DNA material originating from 20
bacterial strains (ATCC MSA-1002, ATCC, USA) was used. Sequences
obtained in this study have been deposited in the European Nucleotide
Archive (ENA) at EMBL-EBI under the accession number PRJEB37267
(\url{https://www.ebi.ac.uk/ena/browser/view/PRJEB37267}).

\hypertarget{sequence-and-data-analysis}{%
\subsubsection{Sequence and data
analysis}\label{sequence-and-data-analysis}}

Obtained sequences were analysed on the computer cluster Isabella
(University Computing Center, University of Zagreb) using mothur
(version 1.43.0) (Schloss et al., 2009) according to the MiSeq Standard
Operating Procedure (MiSeq SOP; \url{https://mothur.org/wiki/MiSeq_SOP})
(Kozich et al., 2013) and recommendations provided by the Riffomonas
project to enhance data reproducibility
(\url{http://www.riffomonas.org/}). For alignment and classification of
sequences the SILVA SSU Ref NR 99 database (release 138;
\url{http://www.arb-silva.de}) was used (Quast et al., 2013; Yilmaz et
al., 2014). Sequences were clustered into operational taxonomic units
(OTUs) at a similarity level of 97 \si{\percent}.

Pipeline data processing and visualization was done using R (version
3.6.0) (R Core Team, 2019) combined with packages vegan (version 2.5-6)
(Oksanen et al., 2019), tidyverse (version 1.3.0) (Wickham, 2017;
Wickham et al., 2019) and multiple other packages (Neuwirth, 2014; Xie,
2014, 2015, 2019a, 2019b, 2019c; Xie et al., 2018; Allaire et al., 2019;
Wilke, 2019; Zhu, 2019). Observed number of OTUs, Chao1, ACE,
exponential of the Shannon diversity index and Inverse Simpson index
were calculated after normalization to the minimum number of reads per
sample using vegan's function \texttt{rrarefy} to account for different
sequencing depths (Oksanen et al., 2019). Chao1 and ACE estimators were
calculated using vegan's function \texttt{estimateR}, while Shannon and
Inverse Simpson diversity indices were retrieved using vegan's function
\texttt{diversity} (Oksanen et al., 2019). To express both diversity
indices in terms of effective number of OTUs the exponential of the
Shannon diversity index was calculated (Jost, 2006). The proportions of
shared OTUs and communities between samples and community types
(seawater, \emph{C. nodosa} {[}mixed{]}, \emph{C. cylindracea}
{[}mixed{]} and \emph{C. cylindracea} {[}monospecific{]}) were expressed
as the Jaccard's (on presence/absence data) and Bray-Curtis similarity
coefficient, respectively. The coefficients were calculated on the OTU
data table using vegan's function \texttt{vegdist} and converted from
dissimilarities to similarities (Borcard et al., 2011; Legendre and
Legendre, 2012; Oksanen et al., 2019). The Principal Coordinates
Analysis (PCoA) was performed on Bray-Curtis dissimilarities based on
OTU abundances using the function \texttt{cmdscale} (Borcard et al.,
2011; Legendre and Legendre, 2012). Differences between communities were
tested by performing the Analysis of Similarities (ANOSIM) using the
vegan's function \texttt{anosim} and 1000 permutations (Oksanen et al.,
2019), while differences in relative contributions or proportions of
shared OTUs and communities were tested by applying the Mann--Whitney
\emph{U} test using the function \texttt{wilcox.test}. In addition,
differences between community type estimators or indices were tested by
performing the Kruskal-Wallis \emph{H} test (function
\texttt{kruskal.test}) followed by a pairwise comparison using the
Mann-Whitney \emph{U} test (function \texttt{pairwise.wilcox}).
Bonferroni correction was used to address the problem of multiple
comparisons.
